\section{Simple Case : $k=2$}
\label{tree:merging:k=2}

\subsection{ITLB}
\label{tree:merging:k=2:ITLB}

Analogously to what was shown in \ref{tree:sorting}, we can compute the
logarithm of the number of possible solution for this problem, i.e.

\begin{theorem}
The ITLB for the merging problem when $k = 2$ with $\card{S_1} = m, \card{S_2}
= n$ is \BigOmega{\log \binom{m+n}{m}}.
\end{theorem}

\begin{proof}
We have to choose the $m$ among $m+n$ positions in $S'$ for the elements of
$S_1$ and then fill the remaining $(m+n) - m = n$ positions with the elements
of $S_2$. The number of leaves of the decision tree is $\binom{m+n}{m}$ hence
the worst minimal height of the tree is $\log \binom{m+n}{m}$.
\end{proof}

\nb{Giving $\log \binom{m+n}{m}$ in the form of the Stirling's approximation
gives us $(m+n) \log(m+n) - m \log m - n \log n$ which clearly express the
information contained in the sorted sequence $S'$ of $m+n$ elements minus the
information we already have.}


\subsection{Algorithms}
\label{tree:merging:k=2:alg}

Without loss of generality, we consider the case where $m \leq n$. We build an
algorithm whose complexity is \BigO{\log \binom{m+n}{m}}.


We begin by computing the general case expression of the number of comparison
required by our ITLB.


\begin{lemma}
\begin{align*}
\log\binom{m+n}{m} &\leq m \log\frac{m+n}{m} + n \log\frac{m+n}{n}\\
\end{align*}
\end{lemma}

\begin{proof}
\begin{align*}
\log\binom{m+n}{m} &= \log\frac{(m+n)!}{m!n!}\\
&= \log(m+n)! - \log m! - \log n!\\
&\leq (m+n) \log(m+n) - m \log m - n \log n\\
&\leq (m+n) \log(m+n) - m \log m - n \log n + (n \log m - n \log m)\\
&\leq (m+n) \log(m+n) - (m+n) \log m - n (\log n - \log m)\\
&\leq (m+n) \log\frac{m+n}{m} - n (\log n - \log m)\\
&\leq m \log\frac{m+n}{m} + n \log\frac{m+n}{m} - n (\log n - \log m)\\
&\leq m \log\frac{m+n}{m} + n \log\frac{m+n}{\sfrac{m}{m} \cdot n}\\
&\leq m \log\frac{m+n}{m} + n \log\frac{m+n}{n}\\
\end{align*}
\end{proof}



In order to feed our intuition on this expression we will analyze its behaviour
on several corner cases.


\subsubsection{First case\texorpdfstring{ : $m = n$}{}}


The first case we will analyze is the case where $m = n$. This case is similar
to what happens during the merge phase of the merge sort algorithm.

\begin{lemma}
\begin{align*}
m = n &\implies \log\binom{m+n}{m} \leq m + n = 2n\\
\end{align*}
\end{lemma}

\begin{proof}
\begin{align*}
\log\binom{m+n}{m} &\leq m \log\frac{m+n}{m} + n \log\frac{m+n}{n}\\
&\leq m \log\frac{2m}{m} + n \log\frac{2n}{n}\\
&\leq m \log 2 + n \log 2\\
&\leq m + n\\
&\leq 2n\\
\end{align*}
\end{proof}

An algorithm performing \BigO{m} is Sequential Merge,


\subsubsection{Second case\texorpdfstring{ : $m \ll n$}{}}


Now we want to see what the bound becomes when $n$ start to grow relatively to
$m$. In the following proof we will assume that $m$ is negligible before $n$.


\begin{lemma}
\begin{align*}
m \ll n &\implies \log\binom{m+n}{m} \leq m \log\frac{n}{m}\\
\end{align*}
\end{lemma}

\begin{proof}
\begin{align*}
\log\binom{m+n}{m} &\leq m \log\frac{m+n}{m} + n \log\frac{m+n}{n}\\
&\leq m \log\frac{n}{m} + n \log\frac{n}{n}\\
&\leq m \log\frac{n}{m}\\
\end{align*}
\end{proof}


This could be intuitively understood as doing $m$ binary searches on sets of
size $\frac{n}{m}$.



\subsubsection{Third case\texorpdfstring{ : $m \ll n \land m = 1$}{}}

Now observe what happens when $m = 1$.

\begin{lemma}
\begin{align*}
m \ll n \land m = 1 &\implies \log\binom{m+n}{m} \leq \log n\\
\end{align*}
\end{lemma}

\begin{proof}
\begin{align*}
\log\binom{m+n}{m} &\leq m \log\frac{n}{m}\\
&\leq \log n\\
\end{align*}
\end{proof}

An algorithm performing \BigO{\log n} is Binary Search.


In conclusion, we need an algorithm performing \BigO{m \log\frac{n}{m}}.
Ideally this algorithm should compete with Sequential Merge for the case where
$m = n$ and with Binary Search for the case where $m = 1$.


\subsubsection{Hwang-Lin}

From now on we have explored special cases, it would be nice if we could
combine the different solutions to build an algorithm of complexity \BigO{\log
\binom{m+n}{m}} that works for any input.

The idea is that the algorithm should behave like :

\begin{itemize}
\item Sequential Merge in case $\frac{n}{m} = 1$
\item $m$ binary searches on sets of size $\frac{n}{m}$ in case $m \ll n$
\item Binary Search in case $m \ll n \land m = 1$
\end{itemize}

Hwang and Lin give such an algorithm in \cite{DBLP:journals/siamcomp/HwangL72}.

The algorithm proceeds by reading sequentially every $\frac{n}{m}$ elements of
$S_2$. Once the algorithm finds the interval of size $\frac{n}{m}$ where the
first element of $S_1$ is to be found, it applies a binary search algorithm on
this interval. We have the position of the first element.
We now want the position of the second element of $S_1$. The position of this
second element is necessarily after the position we just found. Thus, if we
proceed in the same manner for all the other elements of $S_1$ we will only
iterate over the $m$ intervals once and will at most perform $m
\log\frac{n}{m}$ comparisons inside the intervals, hence we get an algorithm
whose complexity is $c \cdot \text{ITLB} + m - 1 = \BigO{ITLB}$.

\nb{We can check that this algorithm matches the \concept{ITLB} found for our
special cases.}
