\section{Examples of Orders, Sets, and Posets}


Using the following examples we try to remove any doubt one might still have on the
concepts we just introduced. These examples can be found in
\citet*{Stanley:2011:ECV:2124415}.

For all examples, $[n] = \left\{{1, 2, \cdots, n}\right\}$, $\N$ is the set
of natural numbers and $\IP$ is the set of positive numbers.

\begin{example}
\label{ex:poset:def}

\item\label{ex:poset:def:a} Let $n \in \IP$. The set $[n]$ with its usual order,
namely the ordering on natural numbers of this set, forms an $n$-element poset
with the special property that any two elements are comparable. This poset is
denoted $\bm{n}$. Of course $\bm{n}$ and $[n]$ coincide as sets, but we use the
notation $\bm{n}$ to emphasize the order structure.

\item\label{ex:poset:def:b} Let $n \in \N$. We can make the set $2^{[n]}$ of
all subsets of $[n]$ into a poset $B_n$ by defining $\S \leq \T$ in $B_n$ if $\S
\subseteq \T$ as sets. One says that $B_n$ consists of the subsets of $[n]$
``ordered by inclusion''.

\item\label{ex:poset:def:c}  Let $n \in \IP$. The set of all positive integer
divisors of $n$ can be made into a poset $D_n$ in a natural way by defining $i
\leq j$ in $D_n$ if $j$ is divisible by $i$ (denoted $i \mid j$).

\item\label{ex:poset:def:d}  Let $n \in \IP$. We can make the set $\Pi_n$ of all
partitions of $[n]$ into a poset (also denoted $\Pi_n$) by defining $\pi \leq
\sigma$ in $\Pi_n$ if every block of $\pi$ is contained in a block of $\sigma$.
For instance, if $n = 9$ and if $\pi$ has blocks $137, 2, 46, 58, 9$, and
$\sigma$ has blocks $13467, 2589$, then $\pi \leq \sigma$. We then say that
$\pi$ is a \emph{refinement} of $\sigma$ and that $\Pi_n$ consists of the
partitions of $[n]$ ``ordered by refinement''.
\end{example}


In \sref{ex:poset:def:a}, $[n]$ is thus the set, the \emph{ordering on natural
numbers of this set} is the order and \emph{the conjunction of the set together
with the order} is the poset. A poset is thus a relational structure $(\S, R)$
and in this special case $\S = \N$ and $R = \le$. Any two elements from this
poset are comparable since the order of this poset is a (weak) total order. We
say that this order is weak because it is not strict, where a strict order is a
trichotomy. An example of trichotomy would have been $R = <$, because then only
one of $x R y$, $x = y$ and $y R x$ can be true.


\sref{ex:poset:def:b} is a classical example to show that posets do not only
work with numbers but also with more complex objects like sets. However, unlike
$\bm{n}$ (\sref{ex:poset:def:a}), the order of this poset is not a total order.
Indeed, let us consider $\S$ and $\T$ two subsets of $[n]$ that each contain an
element of $[n]$ that is not in the other subsets ($\exists s \in \S, s \notin
\T$ and $\exists t \in \T, t \notin \S$). We observe $\S \nsubseteq \T$ and $\T
\nsubseteq \S$ which means that $\S$ and $\T$ are incomparable, and since a
totally ordered set cannot contain incomparable elements the set $B_n$ is only
partially ordered. This is a good example since we now understand what
\emph{incomparable} can mean for a pair of elements.


When we started to explain \sref{ex:poset:def:b} we said that we were working
with sets of more complex objects, but this is not the reason why we found
incomparable objects. With \sref{ex:poset:def:c} we show that even simple
objects like numbers can be considered incomparable. For this to be true, we
simply need to change the relation we consider. If we take for example $n =
12$, we have $D_n = \left\{{1, 2, 3, 4, 6, 12}\right\}$ and neither $3 \mid 4$
nor $4 \mid 3$.


\sref{ex:poset:def:d} is yet another example of a poset that is not total. With
this example we also see that we could give a more abstract meaning to the
order relation (we describe the relation with words that do not refer to a
mathematical operation).
