\section{History}

The first appearance of this problem dates back to 1976. In
\citet*{fredman:1976}, an algorithm making $\BigO{\log e(\P) + 2n}$ queries was
featured. Fredman applies this algorithm to a problem he attributes to Elwyn
Berlekamp: sorting the set $X + Y = \enum{x + y \st x \in X, y \in Y}$.
However,
since \(\log e(\P)\) can be quite small, this algorithm was not matching the \ITLB
for $\log e(\P)$ sublinear in \(n\). Moreover, the algorithm needs exponential time to
choose the right comparison to perform.

From that point, further research had to be made in order to get an algorithm
needing only $\BigO{\log e(\P)}$ comparisons as well as an algorithm running in
polynomial time.

In \citet*{kahn1984balancing}, it is shown that for any finite poset \(\P\)
there always exists a query of the form \(a \ask{\leq} b\) with \(a,b \in \P\)
such that the fraction of linear extensions in which $a$ is smaller than $b$
lies in the interval $(\sfrac{3}{11}, \sfrac{8}{11})$. This is a relaxation of
the well-known \onethirdtwothird conjecture, a conjecture formulated
independently by Fredman, Linial, and Stanley, see
\citet*{linial:1984}. Note that, a simpler proof yielding weaker
bounds was given by \citet*{kahn1991balancing} and better bounds were later
given by \citet*{brightwell1995balancing}, and \citet*{brightwell1999balanced}.
In fact, the only requirement for a practical use (without proving the
conjecture) was to prove that it was true for an interval $(q, 1-q)$, because
then iteratively choosing such a comparison yields an algorithm that performs
$\BigO{\log e(\P)}$ comparisons (precisely $\log_{(1-q)^{-1}} e(\P)$). However, the
algorithm proposed by \citet*{kahn1984balancing} does not find this
\emph{optimal} query in polynomial time.

For the sake of completeness, we hereunder state the \onethirdtwothird
conjecture,

\begin{conjecture}
In any finite poset $\P$ that is not totally ordered, it is always possible to
find a pair $(a,b)$ of elements of $\P$ such that the number of linear
extensions of $\P$ in which $a$ comes before $b$ is between $\sfrac{1}{3}$ and
$\sfrac{2}{3}$ of the total number of linear extensions of $\P$.
\end{conjecture}

The question of its correctness remains open and the
$\sfrac{1}{3}$ bound would be best possible. \citet*{linial:1984} gives a proof
for width-\(2\) posets. Also, there are results due to
\citet*{peczarski:2006} for some classes of partial orders verifying the
conjecture showing that the proportion of partial orders contained in those
classes approaches $1$ as $n$ grows. We can conclude that the proportion of
$n$-element partial orders obeying the \onethirdtwothird conjecture approaches
$1$ as \(n \to \infty\).

Recent findings in the domain include \citet*{zaguia:2011} and
\citet*{peczarski:2008}.
